taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-04-03 paper/manuscript.Rmd
writing

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Kari Norman"
    affiliation: ucb
  - name: "Jorrit Poelen"
    affiliation: jorritt
  - name: "Scott Chamberlain"
    affiliation: ropensci      
  - name: "Carl Boettiger"
    affiliation: ucb, 1
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: jorrit
    address: "Independent consultant, Oakland, CA, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA"
abstract: |
  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources. While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.  Here, we present `taxadb` R package to address this gap.  In contrast to existing tools, `taxadb` provides the following:
  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hiearchical classification.  
  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.
  3) `taxadb` provides a consistent and intuitive data format
  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accomodiate new authorities.

@@

printtable <- function(df){
  df %>% 
    kableExtra::kable("latex", booktabs=T) %>% 
    kable_styling(full_width = F)
}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)
```

@@

## Returning IDs

@@

```{r echo = FALSE, cache = FALSE}
joined %>% printtable()
```

@@

```{r cache = FALSE, echo = FALSE}
joined %>%  
  select(elton_name, redlist_name, mass, category, id) %>% printtable()
```

The `get_ids()` function in `taxadb` can quickly resolve thousands of species names to unique identifiers.

@@

Notice that one species name could not be resolved to a COL identifier.  Perhaps we can resolve this name against a different authority?  As it has an IUCN classification already, the IUCN naming provider seems a natural place to start.  We can use the `synonyms()` table to attempt to identify any other scientific names by which this species is known:

@@

## Synonyms

Walking synonyms between providers should be done with care.  An accepted name can have many synonyms:

@@

Here we see that some accepted identifiers, such as `COL:43082445`, are known by as many as 456 different synonyms!  `get_ids()` function seamlessly handles the resolution of all synonyms to their accepted name usage identifier, since we can unambiguously map from many-to-one.

@@

Some synonyms may also be associated with more than one unique identifier.  This situation is more difficult, as it is impossible to know which accepted name should be used for such synonyms without more context.  The `get_ids()` function will thus return `NA` for any name that matches more than one identifier, just as it will return `NA` for a name that does not match any identifier at all.  Other functions in `taxadb` allow us to explore these names further.  We have already seen this strategy above

@@

For many applications, it is easier to work with the `by_*` class of functions, that first perform the relevant filtering join in the MonetDBLite database before returning the table in memory.  This means that the `data.frame` objects returned by these functions behave as familiar objects. For instance, we can take the top 5 ambiguous names and see all the taxa to which those names may refer:

```{r}
top_5 <- most_ambiguous %>% 
  head(5) %>% 
  pull(scientificName)

by_name(top_5, "col") %>%
  select(scientificName, acceptedNameUsageID) %>% 
  distinct() %>% 
  mutate(acceptedNameUsage = get_names(acceptedNameUsageID, "col")) %>% 
  head(5)
```

@@

- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
- Catalogue of Life annual Species list

@@

The extent of mismatch between providers can be startling.  Catalogue of Life uses ITIS data as one of it's many input sources, and yet we can quickly calculate how many of the (2018) ITIS accepted species names are not in the Catalogue of Life records:

```{r}
itis_col <-
taxa_tbl("itis") %>%
  filter(taxonomicStatus == "accepted", taxonRank == "species") %>%
  select(scientificName) %>% 
  left_join(taxa_tbl("col"), by="scientificName") %>% 
  filter(is.na(acceptedNameUsageID)) %>% 
  count() %>%
  pull(n)

itis_accepted <- taxa_tbl("itis") %>%
  filter(taxonomicStatus == "accepted", taxonRank == "species") %>% 
  count() %>% pull(n)
```
$`r itis_col`$ of the $`r itis_accepted`$ of *accepted names* in ITIS ($`r itis_col / itis_accepted * 100`$%) are not recognized by the (2018) Catalogue of Life as either accepted names or known synonyms.

>>>>

---
title: "taxadb: A High-Performance Local Taxonomic Database Interface"
author:
  - name: "Kari Norman"
    affiliation: ucb
  - name: "Jorrit Poelen"
    affiliation: jorritt
  - name: "Scott Chamberlain"
    affiliation: ropensci      
  - name: "Carl Boettiger"
    affiliation: ucb, 1
address:
  - code: ucb
    address: "Dept of Environmental Science, Policy, and Management, University of California Berkeley, Berkeley CA 94720-3114, USA"
  - code: jorrit
    address: "Independent consultant, Oakland, CA, USA"
  - code: ropensci
    address: "The rOpenSci Project, University of California Berkeley, Berkeley CA 94720-3114, USA"
abstract: |
  A familiar and growing challenge in ecological and evolutionary research is that of reconciling scientific names of relevant taxa when combining data from separate sources. While this problem is already well understood and numerous naming authorities have been created to address the issue, most researchers lack a fast, consistent, and intuitive way to reconcile taxonomic names.  Here, we present `taxadb` R package to address this gap.  In contrast to existing tools, `taxadb` provides the following:  
  1) `taxadb` accesses established naming authorities to resolve synonyms, IDs, and hiearchical classification.  
  2) `taxadb` creates a local database, managed automatically from within R, that provides fast operations on millions of taxonomic names.  
  3) `taxadb` provides a consistent and intuitive data format.  
  4) `taxadb` is built on a simple, extensible and language agnostic design that can easily accomodiate new authorities.

@@

printtable <- function(df){
  df %>% 
    kableExtra::kable("latex", booktabs=T) %>% 
    kable_styling(full_width = F)
}
knitr::opts_chunk$set(cache=TRUE, message=FALSE, warning=FALSE)

# no sci notation integers pleases
options(scipen=999)

```

@@

## Taxonomic IDs

@@

```{r echo = FALSE, cache = FALSE}
joined %>% tidyr::replace_na(list(category = "-")) %>% printtable() # "NA" looks like a category, "-" is clearer visually
```

@@

```{r cache = FALSE, echo = FALSE}
## Just for pretty-printing
joined %>%  
  tidyr::replace_na(list(category = "-", elton_name = "-", redlist_name = "-")) %>%
  select(elton_name, redlist_name, mass, category, id) %>%
  printtable()
```

This results in many more matches, as different scientific names are recognized by the naming provider (Catalog of Life 2018 in this case), as *synonyms* for the same species, and thus resolve to the same taxonomic identifier.  While we have focused on a small example for visual clarity here, the `get_ids()` function in `taxadb` can quickly resolve hundreds of thousands of species names to unique identifiers, thanks to the performance of fast joins in a local MonetDBLite database.

@@

## Unresolved names

Notice that one species name could not be resolved to a COL identifier.  This could indicate a data entry problem such as a misspelling of a name, but failure to resolve properly specified scientific names is not uncommon, as even large scale efforts like the Catalogue of Life are far from a comprehensive list of all published scientific names.

Perhaps we can resolve this name against a different authority?  As it has an IUCN classification already, the IUCN naming provider seems a natural place to start.  We can use the `synonyms()` table to attempt to identify any other scientific names by which this species is known:

@@

## Additional taxonomic information

## Caution using synonyms

This process of walking synonyms between different providers should always be done with care.  Synonyms can arise for many reasons (see [@something] for discussion), and treating accepted names and known synonyms as equivalent or interchangable can quickly lead to innaccurate conclusions.

An accepted name can have many synonyms:

@@

Here we see that some accepted identifiers, such as `COL:43082445`, are known by as many as 456 different synonyms!  `get_ids()` function seamlessly handles the resolution of all these synonyms to their accepted name usage identifier when they can be mapped to a single accepted name.

@@

Some synonyms may also be associated with more than one unique identifier.  This situation is more difficult, as it is impossible to know which accepted name should be used for such synonyms without more context.  The `get_ids()` function will thus return `NA` for any name that matches more than one identifier, just as it will return `NA` for a name that does not match any identifier at all.  Other functions in `taxadb` allow us to explore these names further.

@@

@@

- the Integrated Taxonomic Information System [ITIS; @itis], original formed to standardize taxonomic name usage across many agencies in the United States federal government, 
- the National Center for Biological Information's (NCBI) Taxonomy database, [@ncbi]
- Catalogue of Life (COL) Annual Species list. (2018 version at present)
- Global Biodiversity Information Facility Taxonomic Backbone [GBIF; @gbif].  
- FishBase [@fishbase] taxonomic names
- SeaLifeBase [@sealifebase] taxonomic names
- Wikidata [@jorrit]
- Open Tree Taxonomy, OTT
- International Union for Conservation of Nature and Natural Resources (IUCN) Red list @iucn

Note that some of these providers are primarily syntheses of other existing naming providers.  For instance, COL is based in part on over 130 other name databases (<http://www.catalogueoflife.org/col/info/databases>), including ITIS, FishBase, and the IUCN Redlist.  However, these syntheses may lag behind the most recent versions of their member databases, which can create surprisingly large mismatches.  Because `taxadb` provides access to the different providers using a common format, we can quickly explore this.  For instance, here we count how many accepted names in ITIS can actually be found in COL (as either accepted names or known synonyms):

```{r}
# how many ITIS accepted species names are in COL?
itis_in_col <-
taxa_tbl("itis") %>%
  filter(taxonomicStatus == "accepted", taxonRank == "species") %>%
  select(scientificName) %>% 
  left_join(taxa_tbl("col"), by="scientificName") %>% 
  filter(is.na(acceptedNameUsageID)) %>% 
  count() %>%
  pull(n)

# out of how many total accepted names in ITIS?
itis_accepted <- taxa_tbl("itis") %>%
  filter(taxonomicStatus == "accepted", taxonRank == "species") %>% 
  count() %>% pull(n)
```
$`r prettyNum(itis_in_col, big.mark = ",")`$ of the $`r prettyNum(itis_accepted, big.mark = ",")`$ of *accepted names* in ITIS ($`r round(itis_in_col / itis_accepted * 100)`$%) are not recognized by the (2018) Catalogue of Life as either accepted names or known synonyms.

Consequently, just because the naming providers such as COL, OTT, and GBIF consume the names in ITIS, NCBI, or IUCN, this does not mean we can rely exclusively on these larger syntheses.

<!--
It is important to distinguish between two broad kinds of naming providers.  Efforts such as ITIS, NCBI, FishBase, and SeaLifeBase are largely independent naming assemblies. GBIF and OTT are synthetic 'taxonomic backbones,' created by reconciling species names from a wide range of providers to create a taxonomic backbone utilized internally for their respective efforts.  Catalogue of Life (COL)
-->

@@
